---
output: pdf_document
---

# Analysis

## ETF Price and Return Trends


```{r time_plot_returns, fig.cap=figures("time_plot_returns")}
panel.time.plots <- function (...) {
	lines(...)
	grid()
	abline(h = 0)
}

plot(ret.z, xlab = "Time", panel = panel.time.plots, col = asset.colors, main = "")
figures.add(name = "time_plot_returns", caption = "Timeplot of ETF continuously compounded returns")
```


```{r time_plot_price, fig.cap=figures("time_plot_price")}
panel.time.plots.dates <- function (...) {
	abline(v = dates.macro, col = dates.macro.colors, lty = 2)
	panel.time.plots(...)
}


plot(prices, xlab = "Time", panel = panel.time.plots.dates, col = asset.colors, main = "")
figures.add(name = "time_plot_price", caption = "Timeplot of ETF prices")
```


Analyzing the price data and continuously compounded returns of each of the ETFs in `r figures.cite("time_plot_price")` and `r figures.cite("time_plot_returns")`, it is clear that VFINX and VBISX have the most stable stream of returns, which is reflected in the steady increase in their respective prices over time.

In the case of VFINX, the constant positive progression of the price is to be expected as the stock market has displayed an above-average yearly growth rate over the time horizon considered in the graph. This above-average growth rate is explained by renewed investor confidence in the market in the aftermath of the financial crisis of 2009. Considering the VBISX ETF, which tracks the prices of diversified short-term debt obligations, the price increase can be attributed to a constant reduction in interest rates (yield) of short-term debt in the last five years.^[@ShortTermInterestRates:OECD:2016] The reducing interest rates has a positive effect on the net present value (i.e. price) of the debt, which is reflected in the rising price of the VBISX ETF.

In addition to the price trends inherent to each security, the time plots also indicate that there are significant fluctuations in the prices common to all of the equity ETFs (i.e. except VBLTX and VBISX, which are debt ETFs). These common fluctuations are triggered by specific macroeconomic events, and illustrate congential non-diversifiable risk (i.e. market risk) in each of the equity ETFs.

During the middle months of 2012, many European Union (EU) nations (Greece, Spain, Ireland, etc.) were in serious risk of defaulting (or in the case of Greece, already defaulted) on their debt obligations.^[@EuropePanicSelling:Guardian:2012] This caused a significant reduction in investor confidence in capital markets across the globe, which eventually led to massive selloffs in markets across the world, with investors shifting capital from equity investments to perceptively safer and less-risky debt investments.^[@LowTreasuryNoteYield:CNN:2012] These macroeconomic events are the cause of the simlultaneous significant reduction in equity ETF prices that track multiple regions across the world. This loss in confidence affected markets that have perceptively higher levels of risk compared to others, as is reflected by the sharper drop in emerging markets (VEIEX) when compared to less risky markets such as that of the U.S. (VBLTX). As investors shifted capital to less-risky investments, the surge in invested capital in bond markets caused a drop in bond yields, which was seen in both short-term and long-term bonds. This reduction in bond yields is reflected in the rise in the price of debt ETFs VBLTX and VBISX, as bond yields correlate negatively with the present value (i.e. price) of bonds.

Consider the stock market correction in August 2011, which rattled equity markets worldwide. Investor confidence was damaged when Greece became the first advanced economy to ever default on an International Monetary Fund (IMF) loan payment in July.^[@GreeceIMFDefault:WSJ:2015] This was followed by a surprise devaluation of the Chinese Yuan by the Government of China in August,^[@ChinaYuanDevaluation:Bloomberg:2015] which led to a large drop in the value of companies that use the Chinese Yuan as their primary currency due to their high exposure to foreign exchange risk. Significantly lowered investor confidence, coupled with indicators that the market was inflated led to an eventual global sell-off and inevitable stock market correction in August 2015.^[@BlackMonday:NYT:2015] This is reflected in the sharp drop in prices of ETFs that track regions across the globe. As with the EU credit crisis of 2012, markets that traditionally display higher levels of risk experienced the largest declines in price, as reflected by steeper drops in Emerging and Pacific Market ETFs (VEIEX and VPACX) compared to American and European Market ETFs (VFINX and VEURX).

The presence of market-wide trends is extremely important in the construction of portfolios that comprise ETFs that inherently exhibit these trends. Logically, it can be deduced that trends in the price of a specific ETF that are not common across all other ETFs are attributable to some characteristic of that particular fund, which are fundamentally different from trends that emanate across all of the funds in the market.

Simply put, price trends and fluctuations are illustrations of risk perceived by investors, as a given price reflects the amount of risk an investor is willing to undertake to realize a certain return. This concept can be generalized to the different types of price fluctuations, and would have the following implications:

- Price fluctuations and trends that are specific to a given ETF are an expression of a certain risk factor that is inherent to that particular ETF
- Price fluctuations and trends that are visible across all ETFs are an expression of risk that is perpetuated across the entire market in which the ETFs operate, and is thus not specific to any given ETF

In financial theory, these two types of risk are referred to as diversifiable or unique risk, which is risk that is attributable to some characteristic of a particular ETF, and market or idiosyncratic risk, which is perceived across all ETFs in the market. The main result of this distinction is that combinations of ETFs with particular levels of diversifiable risk can be combined to eliminate their effect, leaving only the idiosyncratic risk of the market, which achieving an enhanced rate of return compared to that of the market as a whole. This is the ideal output of a portfolio, and its realization is the ultimate goal that motivates modern Portfolio Construction Theory.
