---
output: pdf_document
---

# Portfolio Theory

## Introduction

Having conducted a through analysis of statistical parameters related to each individual ETF in the previous sections, we can now consider combinations, or *portfolios* of these ETFs with confidence. To begin, we must first define certain statistics related to each portfolio, and formulas that can be used to calculate the each statistic.

\begin{gather*}
	\text{Let $\mathbf{x}$ be a vector of the asset weights in the portfolio} \\
	\text{Let $\boldsymbol{\mu}$ be a vector of the expected returns of each of the assets} \\
	\text{Let $\boldsymbol{\Sigma}$ be the covariance matrix of the assets} \\
	\\
	\therefore \text{Portfolio Expected Return, } \mu_p = \boldsymbol{\mu} \cdot \mathbf{x} \\
	\therefore \text{Portfolio Variance, } \sigma_p^2 = \mathbf{x}^\intercal \times \boldsymbol{\Sigma} \times \mathbf{x}
\end{gather*}

By using the definitions for portfolio characteristics above, portfolios can be constructed that have desirable characteristics with regard to a certain measurement. For example, a portfolio with a target expected return, but minimal variance can be calculated by performing a constrained optimization of the variance of a portfolio, with the constraint that the expected return of the portfolio is equal to the target expected return. Consider the following portfolios, and the notion of efficient portfolios discussed below.

## Portfolios Allowing Short Sales

### Global Minimum Variance Portfolio

The notion of a minimum variance portfolio is derived from the notion that all of the assets have risks associated with them, and thus the possibility of constructing a portfolio that may yield a level of risk lower than that of any single asset is particularly attractive. This type of portfolio can be constructed by determining the asset weights that would minimize the value of the variance of the portfolio, $\sigma_p$. This problem is described mathematically below:

\begin{gather*}
	\min_{\mathbf{x}} \sigma_p = \mathbf{x}^\intercal \times \boldsymbol{\Sigma} \times \mathbf{x} \\
	\text{subject to } \mathbf{x}^\intercal \cdot \mathbf{1} = 1
\end{gather*}

This constrained optimization can be performed quickly and efficiently on a computer, and the results are displayed below.

```{r port_shorts_minvar_weights, fig.cap = figures("port_shorts_minvar_weights")}
figures.add(name = "port_shorts_minvar_weights", caption = "Global Minimum Variance Portfolio Asset Weights allowing Short Sales")
port.weights.plot(port.shorts.minvar$weights)

```

Displayed in `r figures.cite("port_shorts_minvar_weights")` is a histogram of the asset weights that would construct a portfolio with minimal variance, given the initial assets, thereby satisfying the constrained optimization described above. While the asset weights alone are not useful, statistical parameters regarding the portfolio can be calculated - in particular, the expected return and standard deviation can be calculated. Using this, we can calculate the Sharpe Ratio of the portfolio, as well as 1% and 5% VaRs that would be offered by this portfolio (assuming an initial investment of \$`r formatC(w0, format = "d", big.mark = ",")`).

```{r port_shorts_minvar_table}
tables.add(name = "port_shorts_minvar_stats", caption = "Descriptive statistics for the Global Minimum Variance Portfolio allowing Short Sales")
kable(port.shorts.minvar.table, digits = 6, caption = tables("port_shorts_minvar_stats"))
```

`r tables.cite("port_shorts_minvar_stats")` shows both the monthly, and implied annual statistics for the global minimum variance portfolio. Looking at the implied standard deviation of the portfolio, it is clearly lower than that of any of the individual assets, immediately illustrating the possible benefits of diversification that are offered in this portfolio. Interestingly, this does not cause the asset to have a significantly rate of return, as it still has a higher expected return that that of the worst-performing ETF, VEIEX.

Furthermore, the VaR of this portfolio clearly quantifies the benefit that can be gained from effective diversification. Both the 1% and 5% VaR for this portfolio reflect the extremely low risk it provides. Considering these VaRs relative to the individual asset VaRs, they are significantly lower than the lowest VaRs, offered by VBISX at both time horizons and risk levels.


### Maximum Expected Return Equivalent Efficient Portfolio

An efficient portfolio is defined as a portfolio that can deliver the highest expected return, at a given level of risk. Inverting this statement, it also means that it is any portfolio that provides a target return at the lowest possible risk. To illustrate this, we will calculate an efficient portfolio that has an expected return equal to highest return out of the ETFs, VFINX. The constrained optimization can thus be set up as follows:

\begin{gather*}
	\min_{\mathbf{x}} \sigma_p = \mathbf{x}^\intercal \times \boldsymbol{\Sigma} \times \mathbf{x} \\
	\text{subject to } \mathbf{x}^\intercal \cdot \mathbf{1} = 1 \\
	\boldsymbol{\mu} \cdot \mathbf{x} = E[\text{VFINX}] = `r asset.univar.stats$mean[1]`
\end{gather*}

```{r port_shorts_evfinx_weights, fig.cap = figures("port_shorts_evfinx_weights")}
figures.add(name = "port_shorts_evfinx_weights", caption = "VFINX Return Equivalent Efficient Portfolio Weights allowing Short Sales")
port.weights.plot(port.shorts.evfinx$weights)
```

```{r port_shorts_evfinx_stats}
tables.add(name = "port_shorts_evfinx_stats", caption = "Descriptive Statistics of VFINX Return Equivalent Efficient Portoflio (EVFINX) and VFINX")
kable(port.shorts.evfinx.table, digits = 6, caption = tables("port_shorts_evfinx_stats"))
```

After completing the optimization, the portfolio weights of the VFINX return equivalent efficient portfolio (hereafter referred to as EVFINX) are represented visually in `r figures.cite("port_shorts_evfinx_weights")`. Similar to with the Global Minimum Variance Portfolio, descriptive statistics for the portfolio are displayed in `r tables.cite("port_shorts_evfinx_stats")`. Also shown are the (repeated) descriptive statistics for VFINX. Considering the expected return, they are exactly equal, as required by the conditions of the initial optimization by which the portfolio was calculated. Perhaps more striking however, is the significantly lower standard deviation. By diversifying, an investor could achieve the same expected return as VFINX, but at nearly a quarter of the risk. This effect of this massive reduction in risk is reflected in the extremely high Sharpe ratio of this portfolio compared to that of VFINX, quantifying the significantly higher return per unit of risk. The effect of the lower standard deviation can be quantified to dollars by analyzing the VaRs of both EVFINX and VFINX, as EVFINX has significantly lower VaR at both 1% and 5% compared to that of VFINX.


### Tangency Portfolio

Having displayed the optimization capabilities in terms of both the expected return, and standard deviation, we can now address the Sharpe Ratio. Recall, the Sharpe ratio measures the amount of return that an asset can provide, per unit of risk. Thus, by computing the portfolio that offers the best Sharpe Ratio, we can determine the best possible portfolio to own, given the set of assets. Described below is the constrained optimization that would be performed to compute this portfolio.

\begin{gather*}
	\min_{\mathbf{x}} SR_p = \frac{\mathbf{x}^\intercal \cdot \boldsymbol{\mu} - r_f}{\left( \mathbf{x}^\intercal \times \boldsymbol{\Sigma} \times \mathbf{x} \right)^{\frac{1}{2}}} \\
	\text{subject to } \mathbf{x}^\intercal \cdot \mathbf{1} = 1
\end{gather*}

```{r port_shorts_tangency_weights, fig.cap = figures("port_shorts_tangency_weights")}
figures.add(name = "port_shorts_tangency_weights", caption = "Tangency Portfolio Asset Weights allowing Short Sales")
port.weights.plot(port.shorts.tangency$weights)

```


The resulting asset weights from this optimization can be seen in `r figures.cite("port_shorts_tangency_weights")`. Also, `r tables.cite("port_shorts_tangency_stats")` displays both monthly and annualized descriptive statistics calculated for the tangency portfolio. Looking at the table, both the monthly and annualized Sharpe ratios of this portfolio are significantly higher than any of the assets, with it being nearly three times higher the highest one displayed by any of the ETFs (VBLTX, SR `r asset.univar.sr[4]`).

```{r port_shorts_tangency_stats}
tables.add(name = "port_shorts_tangency_stats", caption = "Descriptive Statistics for the Tangency Portfolio allowing Short Sales (STAN)")
kable(port.shorts.tangency.table, digits = 6, caption = tables("port_shorts_tangency_stats"))
```

Additionally, by combining the tangency portfolio with the risk free asset, it is possible to construct a new compound portfolio that has a better rate of return compared to any other efficient portfolio at any level of risk. Due to the fact that the risk free asset has a standard deviation of 0, the Sharpe ratio of all of the portfolios that lie on this line all have the same optimal Sharpe ratio as the tangency portfolio. This set of compound mean-variant efficient portfolios are displayed with the Markowitz bullet in `r figures.cite("port_shorts_markowitz_bullet")`.

### Markowitz Bullet

```{r port_shorts_markowitz_bullet, fig.cap = figures("port_shorts_markowitz_bullet")}
figures.add(name = "port_shorts_markowitz_bullet", caption = "Risk-Return Graph with the Markowitz Bullet and Mean-Variant Efficient Portfolios allowing Short Sales")
# Plotting Markowitz Bullet
plot.risk.return()
lines(x = port.shorts.efficient.sd, y = port.shorts.efficient.er, pch = 19, type = "o", cex = 0.5, col = port.shorts.efficient.color)
points(x = c(port.shorts.minvar$sd, port.shorts.evfinx$sd, port.shorts.tangency$sd), y = c(port.shorts.minvar$er, port.shorts.evfinx$er, port.shorts.tangency$er), col = port.colors, pch = 4, cex = 2, lwd = 1.5)
abline(h = port.shorts.evfinx$er, col = port.colors[2], lty = 2)
legend(x = "topright", legend = c(asset.names, "SMINVAR Port", "EVFINX Port", "STAN Port"), col = c(asset.colors, port.colors), cex = 0.8, pch = 4)
lines(x = x, y = y, col = tancency.set.color)
par(default.par)
```

The set of efficient portfolios, across all levels of risk is called the efficient frontier. Due to the nature of the shape produced by these portfolios on the risk-return graph, it is commonly referred to as the *Markowitz Bullet*. Mathematically the asset weights of any portfolio that lies on the efficient frontier are an affine combination of any other two portfolios that lie on the line, and thus the set of efficient portfolios can be computed very easily.

`r figures.cite("port_shorts_markowitz_bullet")` displays the efficient frontier on the risk-return graph, along with each of the ETFs themselves, and the other portfolios calculated above. Notice that the set of efficient portfolios provide a higher rate of return at any given level of risk, compared to the individual ETFs.


## Note on Short Sales

The *short sale* of assets refers to the process of borrowing an asset from a broker and selling it on the open market, thereby making you indebted to your broker. To close the trade, one would simply purchase the asset again on the open market and return it to the broker. This process allows investors to benefit from declining prices, as an investor would make a profit if they can buy the asset back later for a price lower than what they sold it for.

This is why the portfolios above have negative weights. The negative weight represents a short position in an asset, and the money that is gained from selling that asset is in turn invested in another. However, due to the implications of being indebted to a broker, and due to the regulatory burdens placed on 401(k) investments, the short sale of assets is not allowed. This means that the portfolios computed above are not suitable 401(k) portfolios due to the fact that they all involve the short sale of some assets. To address this, an analysis of portfolios that do not allow the short sale of assets is conducted below.


## Portfolios Not Allowing Short Sales

### Global Minimum Variance Portfolio

Similar to the portfolio computed above, an optimal portfolio that does not allow the short sale of assets can be constructed. The only difference this time would be that an additional constraint will have to be imposed, restricting the weights of the assets to be greater than 0. This constrained optimization is described below:

\begin{gather*}
	\min_{\mathbf{x}} \sigma_p = \mathbf{x}^\intercal \times \boldsymbol{\Sigma} \times \mathbf{x} \\
	\text{subject to } \mathbf{x}^\intercal \cdot \mathbf{1} = 1 \\
	\mathbf{x_i} > 0 , \,\, \forall \, i
\end{gather*}


```{r port_long_minvar_weights, fig.cap = figures("port_long_minvar_weights")}
figures.add(name = "port_long_minvar_weights", caption = "Global Minimum Variance Portfolio Asset Weights not allowing Short Sales")
port.weights.plot(port.long.minvar$weights)
```

The asset weights of the resulting portfolio are displayed in `r figures.cite("port_long_minvar_weights")`. Additionally, portfolio statistics were computed for this portfolio and are displayed in `r tables.cite("port_long_minvar_stats")`. It is immediately apparent that when compared to the minimum variance portfolio that allows the short sale of assets that the restriction of not being able to short-sell severely impacts the performance of the portfolio.

In addition to having significantly higher risk, there is almost no benefit of using this portfolio, as it is nearly almost composed of a single asset, VBLTX and has nearly identical expected return, which would be well within the bounds of estimation error for that asset. It also has a lower Sharpe ratio compared to the portfolio allowing short sales, which confirms that it is not as risk-efficient. The loss of the ability to short sell assets is particularly pronounced in the levels of VaR of each of the portfolios, as the no-shorts minimum variance portfolio has a VaR at risk that is sometimes four times that of its short-selling counterpart.

```{r port_long_minvar_stats}
tables.add(name = "port_long_minvar_stats", caption = "Descriptive Statistics for the Global Minimum Variance Portfolio not allowing Short Sales (NSMINVAR)")
kable(port.long.minvar.table, digits = 6, caption = tables("port_long_minvar_stats"))
```

### Tangency Portfolio

As with the short sales scenario, a tangency portfolio that maximizes the the Sharpe ratio can be computed, with the restriction of not being allowed to short sell assets. The optimization that would have to be completed is the same as the scenario that allows short sales, again with the exemption that asset weights cannot be negative.

\begin{gather*}
	\min_{\mathbf{x}} SR_p = \frac{\mathbf{x}^\intercal \cdot \boldsymbol{\mu} - r_f}{\left( \mathbf{x}^\intercal \times \boldsymbol{\Sigma} \times \mathbf{x} \right)^{\frac{1}{2}}} \\
	\text{subject to } \mathbf{x}^\intercal \cdot \mathbf{1} = 1 \\
	\mathbf{x_i} > 0 , \,\, \forall \, i
\end{gather*}

```{r port_long_tangency_weights, fig.cap = figures("port_long_tangency_weights")}
figures.add(name = "port_long_tangency_weights", caption = "Tangency Portfolio Asset Weights not allowing Short Sales")
port.weights.plot(port.long.tangency$weights)
```

```{r port_long_tangency_stats}
tables.add(name = "port_long_tangency_stats", caption = "Descriptive Statistics for the Global Minimum Variance Portoflio not allowing Short Sales (NSTAN)")
kable(port.long.tangency.table, digits = 6, caption = tables("port_long_tangency_stats"))
```

Displayed in `r figures.cite("port_long_tangency_weights")` are the asset weights that would create a mean-variance optimized efficient portfolio. Additionally, descriptive statistics for the computed portfolio were calculated and are displayed in `r tables.cite("port_long_tangency_stats")`. Analyzing these statistics, it is immediately clear that the benefit of short selling assets has a large impact on the performance of a mean-variance efficient portfolio. In this case, despite the fact that the standard deviation of the tangency portfolio with no shorts is lower, this is offset by a large gap in the expected returns of the tangency portfolios.

This large disparity is reflected in the extremely large difference in the Sharpe ratios of each of the portfolios. As the tangency portfolio is designed to be optimized to deliver the maximum possible Sharpe ratio, the fact that the lack of short selling causes the Sharpe ratio to be so near that of one of the single assets that comprise the portfolio (VBLTX, SR = `r asset.univar.stats$sd[4]`) is particularly striking.


### Markowitz Bullet

As with the portfolios that allowed short sales, the efficient frontier, which is the set of portfolios that can deliver the maximum possible return at any level of risk can be determined. However, as there is a inequality constraint applied to these portfolios, the same rule of affine combination does not hold. Thus, the efficient frontier must be determined using a *brute force* approach, which entails computing efficient portfolios with the constraint of no short sales applied to a set of expected returns.

```{r port_long_markowitz, fig.cap = figures("port_long_markowitz")}
figures.add(name = "port_long_markowitz", caption = "Risk-Return Graph with the Efficient Frontiers of Portfolios allowing Short Sales and those without Short Sales")
plot.risk.return()
lines(x = port.long.efficient.sd, y = port.long.efficient.er, pch = 19, type = "l", cex = 0.5, col = port.long.efficient.color)
lines(x = port.shorts.efficient.sd, y = port.shorts.efficient.er, pch = 19, type = "o", cex = 0.5, col = port.shorts.efficient.color)
abline(v = 0.02, col = "springgreen", lty = 2)
points(x = c(port.long.minvar$sd, port.long.tangency$sd), y = c(port.long.minvar$er, port.long.tangency$er), col = port.colors[1:2], pch = 4, cex = 2, lwd = 1.5)
legend(x = "topright", legend = c(asset.names, "NSMINVAR Port", "NSTAN Port"), col = c(asset.colors, port.colors[1:2]), cex = 0.8, pch = 4)
```

`r figures.cite("port_long_markowitz")` illustrates the efficient frontier of portfolios that do not allow short sales, alongside the efficient frontier of portfolios that allow short sales. The position of the efficient frontier allowing short sales completely encompasses the frontier that restricts short sales, indicating that it performs better (i.e. offers higher return per unit of risk) at every level of expected return.

To illustrate this further, consider the green line, which represents the 0.02% level of risk. At this level, an efficient portfolio disallowing short sales would yield an expected return of approximately 0.008, while an efficient portfolio allowing short sales at the same level of risk would yield an expected return of 0.014. At this level of risk, the cost of not being able to short sell assets is approximately 0.006 (0.6%) continuously compounded per month, which has a high impact on potential capital gains in the long run.
